Type Error Telemetry at Scale

### 2023-09-30 programming23/paper.tex
mitesh comments

\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\FILL}{\textbf{FILL}}
\newcommand{\dotscale}[1]{\scalebox{0.72}{#1}}
\newcommand{\wideas}[2]{\makebox[\widthof{#2}][l]{#1}}
\newcommand{\twoline}[2]{\parbox[s]{1.4cm}{\flushleft#1\newline#2}}
\newcommand{\chkYes}{\dotscale{\CIRCLE}}
\newcommand{\chkMaybe}{\wideas{\dotscale{\Circle}}{\chkYes}}
\newcommand{\chkNo}{\wideas{}{\chkYes}}
\newcommand{\pct}[1]{\SI{#1}{\percent}}
\newcommand{\percentile}[1]{{#1}th~percentile}
\newcommand{\modefont}[1]{\texttt{#1}}
\newcommand{\mnocheck}{\modefont{nocheck}}
\newcommand{\mnonstrict}{\modefont{nonstrict}}
\newcommand{\mstrict}{\modefont{strict}}
\newcommand{\FS}{background}
\newcommand{\zerowidth}[1]{\makebox[0pt][l]{#1}}
\newcommand{\zeroheight}[1]{\raisebox{0pt}[0pt][0pt]{\(#1\)}}
\newcommand{\nK}[1]{#1\textsc{k}}
\newcommand{\stddev}[1]{\nK{#1}}
\newcommand{\gcell}[1]{\cellcolor{green!20}#1}
\newcommand{\ycell}[1]{\cellcolor{yellow!18}#1}
\newcommand{\ocell}[1]{\cellcolor{orange!29}#1}
\newcommand{\rcell}[1]{\cellcolor{red!30}$\!\!$#1$\!\!$}
\newcommand{\gbox}[1]{\colorbox{green!20}{#1}}
\newcommand{\ybox}[1]{\colorbox{yellow!18}{#1}}
\newcommand{\obox}[1]{\colorbox{orange!29}{#1}}
\newcommand{\rbox}[1]{\colorbox{red!30}{#1}}
\newcommand{\tefsfont}[1]{\textsc{#1}}
\newcommand{\tekey}{\tefsfont{te}}
\newcommand{\fskey}{\tefsfont{fs}}
\newcommand{\QALAN}{\FILL{} Alan:}
\newcommand{\panon}{pseudonymized}
\newcommand{\symbox}[1]{{\makebox[0.8em][c]{{#1}}}}
\newcommand{\addsym}{\symbox{\(\uparrow\)}}
\newcommand{\keepsym}{\symbox{\(=\)}}
\newcommand{\dropsym}{\symbox{\(\downarrow\)}}
% \newcommand{\dsymbox}[1]{\makebox[0.8em][c]{#1}}
% \newcommand{\dropsym}{\dsymbox{\raisebox{0ex}[0.9em]{\raisebox{0.5ex}{\vrule width 0.6em height 0.6pt}}}}

@@

Both modes report the \code{UnknownProperty} error because
misspellings of property names are common enough to report in both
\mstrict{} and \mnonstrict{} mode. See~\cite{bfj-hatra-2021}
for a more detailed discussion of the rationale for \mstrict{} and
\mnonstrict{} mode.

@@

\begin{table}[t]\centering
  \caption{Size of analyzed code: number of files, number of lines, and lines in edit range}
  \label{t:codebase-size}
  % ;; (list* (max* vv) (median < vv) mm (stddev/mean mm vv) (percentile* vv)))
  % ;; percentile* = 0.95 -- 0.99
  % #hash((editrange . (1156036 926 3007043/817 30975.40553032358 (0.95 8220) (0.96 9858) (0.97 13656) (0.98 18956) (0.99 34725)))
  %       (event-count . (6079 138 38689/135 582.5752201972966 (0.95 960) (0.96 1174) (0.97 1304) (0.98 1836) (0.99 3302)))
  %       (files . (54884 7678 55257996/4639 11853.650098459446 (0.95 40029) (0.96 42771) (0.97 45417) (0.98 48688) (0.99 51761)))
  %       (lines . (1089963 3115 38928947/5992 22364.75846303099 (0.95 18561) (0.96 25892) (0.97 27014) (0.98 29725) (0.99 50547)))
  %       (timespan . (1387897376 845648 591343604248/185705 15724745.722577972 (0.95 10696064) (0.96 12715468) (0.97 15596940) (0.98 21153924) (0.99 35450460))))
  %% NOTE: 3 records have +1M lines, all from one nocheck session, all with 3 files
  %% - session "30,762,216,447,324"
  %% NOTE: 133 records have +52K files, all from one nocheck session
  %% - session "330,172,870,224,081" 
  %% (2nd biggest has 41K files "31,650,473,913,529")
%% Files has HUGE numbers thanks to data assets. Forget it.
% Files  &              &
% 11,911 & \stddev{12}  &
%  7,678 &              &
% 51,761 &              &
  \begin{tabular}[t]{lrrrrl}
    ~          & Mean  & Stddev & Median & \pct{99} Pct. & Distribution \\
    Lines      & 6,497 & \stddev{22} & 3,115 & 50,547 & \includegraphics[width=0.14\columnwidth,valign=M]{img/lines-distribution.pdf} \\
    Edit Range & 3,680 & \stddev{31} & 926   & 34,725 & \includegraphics[width=0.14\columnwidth,valign=M]{img/editrange-distribution.pdf}
    %% TODO x-axis starts at 1, not zero
    %% TODO much bigger font or much smaller plot!!!
  \end{tabular}
\end{table}

\Cref{t:codebase-size} summarizes the size of projects in the dataset.
The middle four columns report the mean, standard deviation, median,
and \percentile{99} for the number of lines
and lines in edit range.\footnote{The data also contains the number of files in
each project, but these numbers are difficult to relate to typechecking because
projects may contain hundreds of files that define data assets.  For the
record, the median file count is 7,678 and the 99th percentile is 51,761 files.
}
The plots on the right show zoomed-in distributions of the line and edit range
sizes.
For example, the x-axis of the first plot counts up to 1,000 lines (not showing
a long tail to the right) and the y-axis counts up to 1,000 telemetry records
(not showing a tall spike on the left).

@@

\begin{table}[t]\centering
  \caption{Session size in seconds and in number of records.}
  \label{t:session-size}
  %% NOTE max time = 1 million seconds = 15 days ...  +100 are over 2 days
  \begin{tabular}{lrrrrl} \\
    ~               & Mean & Stddev & Median & \pct{99} Pct. & Distribution \\
    Time Span (sec) & 3,184 & \stddev{16} & 845 & 35,450 & \includegraphics[width=0.14\columnwidth,valign=M]{img/timespan-distribution.pdf} \\
    Record Count    & 286 & 583         & 138 & 3,302  & \includegraphics[width=0.14\columnwidth,valign=M]{img/event-count-distribution.pdf}
    %% TODO x-axis starts at 1
    %% TODO huge font, or smaller plot
  \end{tabular}
\end{table}

@@

On the object level, the analysis offers several lessons for the Luau
developers and other maintainers of gradually-typed languages:
\begin{itemize}
  \item
    Nearly all sessions (\pct{90}) use the default ``no check''
    typechecking mode~(\cref{f:dataset-overview}).
    Since the next typing mode (\mnonstrict{}) requires no annotations and is
    very conservative about the errors it reports, we conjecture that
    developers would have no objections to using it and are simply unaware that
    other modes exist.
%  \item
%    \pct{50} of all errors are in current module, so probably errors appear locally (not in a distant module);
%    \pct{5} of (current? old??) type errors overlap with edit range, so they don't stick around.
  \item
    Sessions tend to fix errors immediately rather than dealing with them in a batch
    or ignoring the errors altogether~(\cref{f:error-density,f:error-changes}).
    % Changes to the number of errors balance one another~(\cref{f:error-density});
    % edits are equally likely to add or remove errors~(\cref{f:error-changes}).
  \item
    All Luau scripts interact with data assets, but no typechecking mode
    has a precise view of the data.
    Improving precision is an important direction for future work.
    Until then, \mstrict{} mode clearly needs to relax its data model
    analysis because it raises even more errors than \FS{} checking~(\cref{f:tefs-compass}).
\end{itemize}

>>>>

\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\FILL}{\textbf{FILL}}
\newcommand{\dotscale}[1]{\scalebox{0.72}{#1}}
\newcommand{\wideas}[2]{\makebox[\widthof{#2}][l]{#1}}
\newcommand{\twoline}[2]{\parbox[s]{1.4cm}{\flushleft#1\newline#2}}
\newcommand{\chkYes}{\dotscale{\CIRCLE}}
\newcommand{\chkMaybe}{\wideas{\dotscale{\Circle}}{\chkYes}}
\newcommand{\chkNo}{\wideas{}{\chkYes}}
\newcommand{\pct}[1]{\SI{#1}{\percent}}
\newcommand{\percentile}[1]{{#1}th~percentile}
\newcommand{\ptile}[1]{\textsc{p}{#1}}
\newcommand{\modefont}[1]{\texttt{#1}}
\newcommand{\mnocheck}{\modefont{nocheck}}
\newcommand{\mnonstrict}{\modefont{nonstrict}}
\newcommand{\mstrict}{\modefont{strict}}
\newcommand{\FS}{background}
\newcommand{\zerowidth}[1]{\makebox[0pt][l]{#1}}
\newcommand{\zeroheight}[1]{\raisebox{0pt}[0pt][0pt]{\(#1\)}}
\newcommand{\nK}[1]{#1\textsc{k}}
\newcommand{\stddev}[1]{\nK{#1}}
\newcommand{\gcell}[1]{\cellcolor{green!20}#1}
\newcommand{\ycell}[1]{\cellcolor{yellow!18}#1}
\newcommand{\ocell}[1]{\cellcolor{orange!29}#1}
\newcommand{\rcell}[1]{\cellcolor{red!30}$\!\!$#1$\!\!$}
\newcommand{\gbox}[1]{\colorbox{green!20}{#1}}
\newcommand{\ybox}[1]{\colorbox{yellow!18}{#1}}
\newcommand{\obox}[1]{\colorbox{orange!29}{#1}}
\newcommand{\rbox}[1]{\colorbox{red!30}{#1}}
\newcommand{\tefsfont}[1]{\textsc{#1}}
\newcommand{\tekey}{\tefsfont{te}}
\newcommand{\fskey}{\tefsfont{fs}}
\newcommand{\QALAN}{\FILL{} Alan:}
\newcommand{\panon}{pseudonymized}
\newcommand{\symbox}[1]{{\makebox[0.8em][c]{{#1}}}}
\newcommand{\addsym}{\symbox{\(\uparrow\)}}
\newcommand{\keepsym}{\symbox{\(=\)}}
\newcommand{\dropsym}{\symbox{\(\downarrow\)}}
% \newcommand{\dsymbox}[1]{\makebox[0.8em][c]{#1}}
% \newcommand{\dropsym}{\dsymbox{\raisebox{0ex}[0.9em]{\raisebox{0.5ex}{\vrule width 0.6em height 0.6pt}}}}

@@

Both modes report the \code{UnknownProperty} error because
misspellings of property names are common enough to report in both
\mstrict{} and \mnonstrict{} mode.
\mnocheck{} reports no error because the program is syntactically valid.
See~\cite{bfj-hatra-2021} for a more detailed discussion of the rationale
behind the type systems.

@@

\begin{table}[t]\centering
  \caption{Size of analyzed code: number of files, number of lines, and lines in edit range}
  \label{t:codebase-size}
  % ;; (list* (max* vv) (median < vv) mm (stddev/mean mm vv) (percentile* vv)))
  % ;; percentile* = 0.95 -- 0.99
  % #hash((editrange . (1156036 926 3007043/817 30975.40553032358 (0.95 8220) (0.96 9858) (0.97 13656) (0.98 18956) (0.99 34725)))
  %       (event-count . (6079 138 38689/135 582.5752201972966 (0.95 960) (0.96 1174) (0.97 1304) (0.98 1836) (0.99 3302)))
  %       (files . (54884 7678 55257996/4639 11853.650098459446 (0.95 40029) (0.96 42771) (0.97 45417) (0.98 48688) (0.99 51761)))
  %       (lines . (1089963 3115 38928947/5992 22364.75846303099 (0.95 18561) (0.96 25892) (0.97 27014) (0.98 29725) (0.99 50547)))
  %       (timespan . (1387897376 845648 591343604248/185705 15724745.722577972 (0.95 10696064) (0.96 12715468) (0.97 15596940) (0.98 21153924) (0.99 35450460))))
  %% NOTE: 3 records have +1M lines, all from one nocheck session, all with 3 files
  %% - session "30,762,216,447,324"
  %% NOTE: 133 records have +52K files, all from one nocheck session
  %% - session "330,172,870,224,081" 
  %% (2nd biggest has 41K files "31,650,473,913,529")
%% Files has HUGE numbers thanks to data assets. Forget it.
% Files  &              &
% 11,911 & \stddev{12}  &
%  7,678 &              &
% 51,761 &              &

  %% FILL P99 or P01
  \begin{tabular}[t]{lrrrrl}
    ~          & Mean  & Stddev & Median & \ptile{99} & Distribution \\
    Lines      & 6,497 & \stddev{22} & 3,115 & 50,547 & \includegraphics[width=0.14\columnwidth,valign=M]{img/lines-distribution.pdf} \\
    Edit Range & 3,680 & \stddev{31} & 926   & 34,725 & \includegraphics[width=0.14\columnwidth,valign=M]{img/editrange-distribution.pdf}
    %% TODO x-axis starts at 1, not zero
    %% TODO much bigger font or much smaller plot!!!
  \end{tabular}
\end{table}

\Cref{t:codebase-size} summarizes the size of projects in the dataset.
The middle four columns report the mean, standard deviation, median,
and \percentile{99} for the number of lines
and lines in edit range.\footnote{The data also contains the number of files in
each project, but these numbers are difficult to relate to typechecking because
projects may contain hundreds of files that define data assets.  For the
record, the median file count is 7,678 and the \percentile{99} is 51,761 files.
}
The plots on the right show zoomed-in distributions of the line and edit range
sizes.
For example, the x-axis of the first plot counts up to 1,000 lines (not showing
a long tail to the right) and the y-axis counts up to 1,000 telemetry records
(not showing a tall spike on the left).

@@

\begin{table}[t]\centering
  \caption{Session size in seconds and in number of records.}
  \label{t:session-size}
  %% NOTE max time = 1 million seconds = 15 days ...  +100 are over 2 days
  \begin{tabular}{lrrrrl} \\
    ~               & Mean & Stddev & Median & \ptile{99} & Distribution \\
    Time Span (sec) & 3,184 & \stddev{16} & 845 & 35,450 & \includegraphics[width=0.14\columnwidth,valign=M]{img/timespan-distribution.pdf} \\
    Record Count    & 286 & 583         & 138 & 3,302  & \includegraphics[width=0.14\columnwidth,valign=M]{img/event-count-distribution.pdf}
    %% TODO x-axis starts at 1
    %% TODO huge font, or smaller plot
  \end{tabular}
\end{table}

@@

On the object level, the analysis offers several lessons for the Luau
developers and other maintainers of gradually-typed languages:
\begin{itemize}
  \item
    Nearly all sessions (\pct{90}) use the default ``no check''
    typechecking mode~(\cref{f:dataset-overview}).
    Since the next typing mode (\mnonstrict{}) requires no annotations and is
    very conservative about the errors it reports, we conjecture that
    developers would have no objections to using it and are simply unaware that
    other modes exist.
%  \item
%    \pct{50} of all errors are in current module, so probably errors appear locally (not in a distant module);
%    \pct{5} of (current? old??) type errors overlap with edit range, so they don't stick around.
  \item
    Sessions tend to fix errors immediately rather than dealing with them in a batch
    or ignoring the errors altogether~(\cref{f:error-density,f:error-changes}).
    This should give Roblox confidence to make \mnonstrict{} the default in the future.
    % Changes to the number of errors balance one another~(\cref{f:error-density});
    % edits are equally likely to add or remove errors~(\cref{f:error-changes}).
  \item
    All Luau scripts interact with data assets, but no typechecking mode
    has a precise view of the data.
    Improving precision is an important direction for future work.
    Until then, \mstrict{} mode clearly needs to relax its data model
    analysis because it raises even more errors than \FS{} checking~(\cref{f:tefs-compass}).
\end{itemize}
